---
title: "Analysis of Mean"
author: "Ashley Berniche"
format: html
editor: source
---

Here is some new data to play with a bit to explore one-way and two-way analysis of variance models.  These data represent weight data associated with three different diet types.  There are 76 observations of individuals with the following characteristics:

- Gender (0/1)
- Age (earth years. 🤓)
- Height (cm)
- Pre.weight & weight6weeks (kg)
- Diet (1,2,3)

The data are located as a CSV file named `DietData.csv` linked from the Canvas page.  Load the data in and format things so they look correctly to you.

```{r}
library(readr)
library(dplyr)
library(knitr)



df <- read_csv("DietData.csv")


```
## Exploratory Data Anlsysis

One of the first things to do is to look at the data and see if there are any obvious things.  Go ahead and explore these data visually.  What do you see?

There are 76 observations of 7 variables.All of the people are adults. 

```{r}
View(df)
```



## Estimating Mean Values

Make a table of Weight Loss by gender and diet.

```{r}
summary_table <- df %>%
  group_by(gender, Diet) %>%
  summarise(
    Mean_Weight_Loss = mean(weight6weeks, na.rm = TRUE),
    SD_Weight_Loss = sd(weight6weeks, na.rm = TRUE),
    n = n()
  )


kable(summary_table, caption = "Weight Loss by Gender and Diet") 
```


## 1-Way Analysis of Variance

*Diet Issues:* 


The underlying linear model.


$$
y_{ij} = \mu + \tau_{Diet, i} + \epsilon_j
$$


Test the null hypothesis, $H_O:$ *There is no effect of diet on weight loss* (e.g., $\tau_{Diet-1} = \tau_{Diet-2} = \tau_{Diet-3} = 0.0$).  Is there evidence for one diet producing more weight loss than the others?  Create an `aov` analysis and assign it to the variable `fit.diet` and examine its contents.

```{r}
df$Diet <- as.factor(df$Diet)

fit.diet <- aov(weight6weeks ~ Diet, data = df)


summary(fit.diet)
```

Are they all significantly different?  Try the `TukeyHSD()` Interpret the results.
With a p value of 0.656, we fail to reject the null hypothesis, there is no statistical significance. 

```{r}
TukeyHSD(fit.diet)
```
 How much of the variation is explained?  If you notice when you do a summary from a `lm()` (regression) model, it gives you the $R^2$ values directly (remember $R^2 = \frac{SS_{model}}{SS_{Error}}$).  Does `summary()` of your `aov` model give you that?
 The R squared is 0.01668, meaning only 1.668% of the variation can be explained by diet. The summary of the aov does not give the R squared value. 
 
```{r}
fit.lm <- lm(weight6weeks ~ Diet, data = df)
summary(fit.lm)$r.squared
```

Since I asked the question, the answer is probably no.  Why does it not do this?  Probably for historical reasons, which are a bit of a pain in the backside.  That being said, there are some `tidy` ways to fix this issue.  I'm going to use the `broom` package which allows us to clean up (or tidy if you will) the model objects.  This will take the model object and pull out *all* the ANOVA table stuff and put it into a `tibble`.

```{r}
library(broom)

tidy_diet <- tidy( fit.diet )
tidy_diet
```

Now, since it is all easily accessible, we can calculate the $R^2$ from the new model output.

```{r}
r2_Diet <- tidy_diet$sumsq[1] / sum( tidy_diet$sumsq )
r2_Diet 
```


 

*Gender:* 

The underlying linear model.


$y_{ij} = \mu + \tau_{gender, i} + \epsilon_j$



Independent of the diet, test the null hypothesis $H_O:$ *There is no difference in weight loss between genders* (e.g., $\tau_{gender-0} = \tau_{gender-2} = 0.0 $).  Is there evidence for one gender being significantly different than another?  How much of the variation is explained (another $R^2$ by gender)?

```{r}
df$gender <- as.factor(df$gender)
fit.gender <- aov(weight6weeks - pre.weight ~ gender, data = df)
summary(fit.gender)

aov_summary <- summary(fit.gender)
SS_model <- aov_summary[[1]]["gender", "Sum Sq"]
SS_residuals <- aov_summary[[1]]["Residuals", "Sum Sq"]
SS_total <- SS_model + SS_residuals
R2_gender <- SS_model / SS_total
R2_gender

tidy_gender <- tidy(fit.gender)
tidy_gender



```

How do you interpret these results thus far?

The p value is much greater than 0.05, so the we fail to reject the null hypothesis. Also, with an R squared of 0.00059, only 0.059% of weight loss can be explained by gender, giving it little to no impact on weight loss.

# Do genders respond differently to diets?


$y_{ijk} = \mu + \tau_{Diet,i} + \tau_{gender,j} + \epsilon_k$

So here we do something a bit different.  We want to simultaneously ask the following questions:

- Do diets influence weight loss?
- Do genders influence weight loss?
- Is there an interaction where different genders respond differently to different diets?

In $R$, this is done as:

```{r}
fit.diet <- aov(weight6weeks - pre.weight ~ Diet, data = df)
summary(fit.diet)
```

```{r}
fit.gender <- aov(weight6weeks - pre.weight ~ gender, data = df)
summary(fit.gender)
```

```{r}
fit.interaction <- aov(weight6weeks - pre.weight ~ Diet * gender, data = df)
summary(fit.interaction)

r2_Diet <- tidy_diet$sumsq[1] / sum( tidy_diet$sumsq )
r2_Diet 
```

What is the $R^2$ for this model?

```{r}
aov_summary <- summary(fit.interaction)
SS_model <- sum(aov_summary[[1]]$`Sum Sq`[1:3]) 
SS_residuals <- aov_summary[[1]]$`Sum Sq`[4] 


SS_total <- SS_model + SS_residuals


R2 <- SS_model / SS_total
R2
```


# Which Model is Best?

How would you compare the models you generated?  How do you interpret the findings?

The first models did not find a relationship between diet and gender and weight loss, but the second model found that diet has an effect on weight loss, and that the interaction between gender and diet is significant as well. With an R squared of 0.20, it appears the second model fits the data better. The interaction model is able to detect more of the nuiance of the data and identify relationships between the data that the simpler model misses. 