---
title: "Final"
author: "Ashley Berniche"
format: html
editor: source
---

## Introduction

The Sonora Desert bark beetle, *Araptus attenuatus*, is a beetle species that lives within a specific host plant, *Euphoria lomelli*. These plants, and their beetles, are only found within the desert flora of the Sonora desert. Due to climate change, the range of suitable habitat for the host plant may be impacting the life-history of the bark beetle. While preliminary studies have found that habitat suitability may effect the beetle's sex ratios, more research needs to be done to determine the relationship between sex ratio, habitat, and the landscape changes occuring due to climate change. This study aims to explore the relationship, if any, between these factors.

```{r setup, include=FALSE}


library(tidyverse)  
library(readr)      
library(raster)     
library(sf)         
library(knitr)      
library(kableExtra) 

locations <- read_csv("Arapat_Locations.csv")

samples <- read_csv("Arapat_Samples.csv")

suitability_now_path <- "Suitability_now.tif" 
suitability_now <- raster(suitability_now_path)


suitability_lgm_path <- "Suitability_lgm.asc"  
suitability_lgm <- raster(suitability_lgm_path)


```

## Methods

Research was conducted to sample the host plant and count the number of individual beetles found, along with their sex and phenotype.

Using this data, the sex ratios per site were calculated, An ANOVA test was done to determine if there was a difference in sex ratio among the different sampling locations.

To determine which sites had sex ratios that deviated from equal proportions of males and females, a binomial test was done for each site. Sites with a p-value less than 0.05 were determined to have a deviation from an equal proportion in their sex ratio.

To determine if there is a relationship between habitat suitability and sex ratio, sex ratios for each site were collapsed into a single value. Habitat suitability data was then extracted from the raster and compared to sex ratio data using a regression.

Phenotype was added to the regression to determine of the inclusion of type A or B would increase the relationship among habitat suitability and sex ratio. The regression was compared to just habitat suitability to determine if the model improved.

After exploring the relationship between sex ratio and habitat suitability, a raster containing habitat suitability from the last glacial maximum was used to determine if the availability of suitable habitat was changing. A change in available habitat could be linked to climate change. Change in habitat was calculated and then displayed visually as a bar graph and spatially.

Using the model developed for current conditions, suitability data from the last glacial maximum was used to estimate historical sex ratios, which were then compared to historic habitat suitability.

##Results

The result of the ANOVA test returned a p-value of 0.00307, indicating that there is a difference in sex ratio among sampling locations. 

```{r chunk2, echo=FALSE}


sex_ratios <- samples %>%
  group_by(Site, Plant) %>%
  summarise(
    total_beetles = n(),
    males = sum(Sex == "Male"),
    females = sum(Sex == "Female"),
    sex_ratio = males / total_beetles
  )

site_sex_ratios <- sex_ratios %>%
  group_by(Site) %>%
  summarise(
    mean_sex_ratio = mean(sex_ratio),
    sd_sex_ratio = sd(sex_ratio),
    n = n()
  )

anova_result <- aov(sex_ratio ~ Site, data = sex_ratios)
summary(anova_result)

sex_ratios %>%
  ggplot(aes(x = Site, y = sex_ratio)) +
  geom_boxplot() +
  labs(
    title = "Sex Ratios of Beetles by Site",
    x = "Site",
    y = "Sex Ratio (Proportion of Males)"
  ) +
  theme_minimal()


print("ANOVA Results:")
print(summary(anova_result))

sex_ratio_table <- site_sex_ratios %>%
  # Round the relevant columns to 3 decimal places
  mutate(
    mean_sex_ratio = round(mean_sex_ratio, 3),
    sd_sex_ratio = round(sd_sex_ratio, 3)
  ) %>%
 
  knitr::kable(
    caption = "Summary of Sex Ratios by Site",
    col.names = c("Site", "Mean Sex Ratio", "SD of Sex Ratio", "Number of Plants"),
    align = "c"
  ) %>%
  # Add kable styling options
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )

sex_ratio_table



```

Of the 31 sites sampled, six sites had sex ratios that significantly differed from having equal distribution between male and female.In five of the six sites, males were the dominant sex sampled. 

```{r chunk3, echo=FALSE}

samples <- samples %>%
  mutate(Sex = str_trim(Sex))  

site_sex_counts <- samples %>%
  group_by(Site) %>%
  summarise(
    total_males = sum(Sex == "Male", na.rm = TRUE),
    total_females = sum(Sex == "Female", na.rm = TRUE),
    total_beetles = total_males + total_females
  ) %>%
  mutate(observed_ratio = total_males / total_beetles)

sex_ratio_tests <- site_sex_counts %>%
  rowwise() %>%
  mutate(
    p_value = ifelse(
      total_beetles > 0,
      binom.test(total_males, total_beetles, p = 0.5)$p.value,
      NA_real_
    )
  ) %>%
  ungroup() %>%
  mutate(significant = p_value < 0.05)

sex_ratio_test_table <- sex_ratio_tests %>%
  mutate(
    observed_ratio = round(observed_ratio, 3),
    p_value = round(p_value, 3)
  ) %>%
  knitr::kable(
    caption = "Binomial Test Results for Sex Ratios by Site",
    col.names = c("Site", "Total Males", "Total Females", "Total Beetles", "Observed Ratio", "P-Value", "Significant"),
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )


sex_ratio_test_table



```

Based on the results of the regression, only 2.74% of the variance in sex ratio could be explained by suitability. Additionally, the p-value was 0.3736, indicating that the relationship is not significant. Based on the results of this model, habitat suitability alone is not significant enough to explain sex ratios at the different sites. 

```{r chunk4, echo=FALSE}


arapat_samples <- samples %>%
  mutate(Sex = str_trim(Sex)) 

collapsed_sex_ratios <- arapat_samples %>%
  group_by(Site) %>%
  summarise(
    total_males = sum(Sex == "Male", na.rm = TRUE),
    total_females = sum(Sex == "Female", na.rm = TRUE),
    total_beetles = total_males + total_females,
    observed_ratio = total_males / total_beetles
  ) 

  arapat_locations <- locations %>%
  mutate(Site_number = as.numeric(str_extract(Site, "\\d+"))) 

arapat_locations <- arapat_locations %>%
  arrange(Site_number)

arapat_locations_sf <- st_as_sf(arapat_locations, coords = c("Longitude", "Latitude"), crs = 4326)

overlay_suitability <- raster::extract(suitability_lgm, arapat_locations_sf)
arapat_locations <- arapat_locations %>% mutate(suitability = overlay_suitability)

suitability_sex_ratio <- collapsed_sex_ratios %>%
  left_join(arapat_locations, by = "Site") %>%
  filter(!is.na(suitability))  

model <- lm(observed_ratio ~ suitability, data = suitability_sex_ratio)
regression_summary <- summary(model)

print("Linear Regression Results:")
print(regression_summary)

suitability_sex_ratio %>%
  ggplot(aes(x = suitability, y = observed_ratio)) +
  geom_point(size = 3, color = "blue", alpha = 0.7) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Relationship Between Habitat Suitability and Sex Ratio",
    x = "Habitat Suitability",
    y = "Observed Sex Ratio (Proportion of Males)"
  ) +
  theme_minimal()

site_data_table <- suitability_sex_ratio %>%
  mutate(
    observed_ratio = round(observed_ratio, 3),  # Round observed ratios to 3 decimal places
    suitability = round(suitability, 3)         # Round suitability values to 3 decimal places
  ) %>%
  dplyr::select(Site, observed_ratio, suitability) %>%
  knitr::kable(
    caption = "Summary of Sex Ratios and Habitat Suitability by Site",
    col.names = c("Site", "Observed Sex Ratio", "Habitat Suitability"),
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )

# To display the table in markdown output:
site_data_table


```






```{r chunk5, echo=FALSE}


arapat_locations_sf <- st_as_sf(arapat_locations, coords = c("Longitude", "Latitude"), crs = 4326)

overlay_suitability <- raster::extract(suitability_lgm, arapat_locations_sf)
arapat_locations <- arapat_locations %>% mutate(suitability = overlay_suitability)

raster_df <- as.data.frame(suitability_lgm, xy = TRUE)
colnames(raster_df) <- c("Longitude", "Latitude", "Suitability")


ggplot() +
  geom_raster(data = raster_df, aes(x = Longitude, y = Latitude, fill = Suitability)) +
  scale_fill_gradient(low = "white", high = "red") +  
  geom_sf(data = arapat_locations_sf, color = "black", size = 3) + 
  geom_sf(data = arapat_locations_sf, aes(color = suitability), size = 2) +  
  scale_color_gradient(low = "white", high = "red") +
  labs(
    title = "Habitat Suitability and Location Data",
    x = "Longitude",
    y = "Latitude",
    color = "Habitat Suitability"
  ) +
  theme_minimal() +
  theme(legend.position = "right")



```



```{r}
#| echo: false

merged_data <- left_join(suitability_sex_ratio, samples, by = "Site")


# suitability alone
model_1 <- lm(observed_ratio ~ suitability, data = merged_data)
summary(model_1)

# Phenotype_A, and Phenotype_B
model_2 <- lm(observed_ratio ~ suitability + PhenotypeA + PhenotypeB, data = merged_data)
summary(model_2)

cat("Model 1 R-squared:", summary(model_1)$r.squared, "\n")
cat("Model 1 Adjusted R-squared:", summary(model_1)$adj.r.squared, "\n")
cat("Model 2 R-squared:", summary(model_2)$r.squared, "\n")
cat("Model 2 Adjusted R-squared:", summary(model_2)$adj.r.squared, "\n")

cat("Model 1 AIC:", AIC(model_1), "\n")
cat("Model 2 AIC:", AIC(model_2), "\n")

cat("Model 1 p-values:", summary(model_1)$coefficients[,4], "\n")
cat("Model 2 p-values:", summary(model_2)$coefficients[,4], "\n")

anova(model_1, model_2)


```

The analysis compared two linear regression models to examine the relationship between the observed sex ratio (observed_ratio) and habitat suitability, with and without the inclusion of phenotypic variables (PhenotypeA and PhenotypeB). In the first model, which included only habitat suitability, the coefficient for suitability was statistically significant (p \< 2e-16), indicating a negative relationship with the observed sex ratio. The model explained 2.7% of the variance in the data, with an R-squared of 0.02739 and an Adjusted R-squared of 0.02708. The second model, which added PhenotypeA and PhenotypeB, showed a slight improvement in model fit, with an R-squared of 0.06388 and an Adjusted R-squared of 0.06297. This model revealed that both phenotypes also significantly influenced the observed sex ratio. PhenotypeA had a small positive effect (p = 0.0407), while PhenotypeB had a stronger negative effect (p \< 2e-16). The inclusion of phenotypic variables improved model performance, as reflected in a lower AIC for Model 2 (-7525.526 compared to -7411.002 for Model 1), and a significant improvement in the F-statistic (60.331, p \< 2.2e-16). Despite the improvement, the overall explanatory power of the model remains modest, with the combined predictors explaining approximately 6.4% of the variance in the observed sex ratio.

Using the data from the last glacial maximum and the sampling locations, has the suitability changed at each location (e.g., was it as suitable 20,000 years ago as today)?

```{r}
#| echo: false

suitability_now_values <- raster::extract(suitability_now, arapat_locations_sf)
suitability_lgm_values <- raster::extract(suitability_lgm, arapat_locations_sf)

arapat_locations <- arapat_locations %>%
  mutate(suitability_now = suitability_now_values,
         suitability_lgm = suitability_lgm_values)

arapat_locations <- arapat_locations %>%
  mutate(suitability_change = suitability_now - suitability_lgm)

ggplot(arapat_locations, aes(x = Site, y = suitability_change)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Change in Habitat Suitability from Last Glacial Maximum to Present",
    x = "Site",
    y = "Change in Suitability (Present - LGM)"
  ) +
  theme_minimal()

suitability_now_df <- as.data.frame(suitability_now, xy = TRUE)
colnames(suitability_now_df) <- c("Longitude", "Latitude", "Suitability_Now")

suitability_lgm_df <- as.data.frame(suitability_lgm, xy = TRUE)
colnames(suitability_lgm_df) <- c("Longitude", "Latitude", "Suitability_LGM")

map_now <- ggplot() +
  geom_raster(data = suitability_now_df, aes(x = Longitude, y = Latitude, fill = Suitability_Now)) +
  scale_fill_gradient(low = "white", high = "red", name = "Suitability") +
  geom_sf(data = arapat_locations_sf, color = "black", size = 1) +
  labs(
    title = "Current Habitat Suitability",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()

map_lgm <- ggplot() +
  geom_raster(data = suitability_lgm_df, aes(x = Longitude, y = Latitude, fill = Suitability_LGM)) +
  scale_fill_gradient(low = "white", high = "red", name = "Suitability") +
  geom_sf(data = arapat_locations_sf, color = "black", size = 1) +
  labs(
    title = "Habitat Suitability During Last Glacial Maximum",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()

print(map_now)
print(map_lgm)

```

Predict the distribution of the historical sex ratio by applying the model you developed for current conditions to the suitability estimated from the last glacial maximum. Across the landscape, do you detect any trends that may be due to the differences in climate, as measured by our estimates of habitat suitability?

```{r}
#| echo: false

# Load raster files for habitat suitability
suitability_now <- raster("Suitability_now.tif")  # Current habitat suitability raster
suitability_lgm <- raster("Suitability_lgm.asc")  # LGM habitat suitability raster

# Load site locations data
arapat_locations <- read.csv("Arapat_Locations.csv")  # Ensure it has Site, Latitude, Longitude columns

# Convert locations to an sf object
arapat_locations_sf <- st_as_sf(
  arapat_locations,
  coords = c("Longitude", "Latitude"),
  crs = 4326
)

# Transform CRS of locations to match the rasters
arapat_locations_sf <- st_transform(arapat_locations_sf, crs = crs(suitability_now))

# Convert sf object to SpatialPointsDataFrame for raster compatibility
arapat_locations_sp <- as(arapat_locations_sf, "Spatial")

# Extract habitat suitability values from the rasters
arapat_locations <- arapat_locations %>%
  mutate(
    current_suitability = raster::extract(suitability_now, arapat_locations_sp),
    lgm_suitability = raster::extract(suitability_lgm, arapat_locations_sp)
  )

# Load sample data
arapat_samples <- read.csv("Arapat_Samples.csv")  # Ensure it has Site, Sex, etc.

# Calculate observed sex ratios by site
observed_sex_ratio <- arapat_samples %>%
  group_by(Site) %>%
  summarise(
    total_males = sum(Sex == "Male", na.rm = TRUE),
    total_females = sum(Sex == "Female", na.rm = TRUE),
    observed_ratio = total_males / total_females
  )

# Merge sex ratio data with habitat suitability data
merged_data <- left_join(arapat_locations, observed_sex_ratio, by = "Site")

# Linear regression: Observed sex ratio vs. current suitability
model_current <- lm(observed_ratio ~ current_suitability, data = merged_data)
summary(model_current)

# Predict historical sex ratio using the model and LGM suitability
merged_data <- merged_data %>%
  mutate(predicted_sex_ratio_lgm = predict(model_current, newdata = merged_data %>% mutate(current_suitability = lgm_suitability)))

# Visualization: Predicted historical sex ratio by site
ggplot(merged_data, aes(x = Site, y = predicted_sex_ratio_lgm)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  labs(
    title = "Predicted Historical Sex Ratio During the Last Glacial Maximum",
    x = "Site",
    y = "Predicted Sex Ratio"
  ) +
  theme_minimal()

# Visualization: Predicted historical sex ratio vs. habitat suitability (LGM)
ggplot(merged_data, aes(x = lgm_suitability, y = predicted_sex_ratio_lgm)) +
  geom_point(color = "steelblue") +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Predicted Sex Ratio vs. Habitat Suitability (LGM)",
    x = "Habitat Suitability (LGM)",
    y = "Predicted Sex Ratio"
  ) +
  theme_minimal()


# Output the AIC for model comparison
aic_current <- AIC(model_current)
print(paste("AIC for current suitability model:", aic_current))



```

##Discussion
