---
title: "Final"
author: "Ashley Berniche"
format: html
editor: source
---

## Introduction

The Sonora Desert bark beetle, *Araptus attenuatus*, is a beetle species that lives within a specific host plant, *Euphoria lomelli*. These plants, and their beetles, are only found within the desert flora of the Sonora desert. Due to climate change, the range of suitable habitat for the host plant may be impacting the life-history of the bark beetle. While preliminary studies have found that habitat suitability may effect the beetle's sex ratios, more research needs to be done to determine the relationship between sex ratio, habitat, and the landscape changes occuring due to climate change. This study aims to explore the relationship, if any, between these factors.

```{r setup, include=FALSE}


library(tidyverse)  
library(readr)      
library(raster)     
library(sf)         
library(knitr)      
library(kableExtra) 

locations <- read_csv("Arapat_Locations.csv")

samples <- read_csv("Arapat_Samples.csv")

suitability_now_path <- "Suitability_now.tif" 
suitability_now <- raster(suitability_now_path)


suitability_lgm_path <- "Suitability_lgm.asc"  
suitability_lgm <- raster(suitability_lgm_path)


```

## Methods

Research was conducted to sample the host plant and count the number of individual beetles found, along with their sex and phenotype.

Using this data, the sex ratios per site were calculated, An ANOVA test was done to determine if there was a difference in sex ratio among the different sampling locations.

To determine which sites had sex ratios that deviated from equal proportions of males and females, a binomial test was done for each site. Sites with a p-value less than 0.05 were determined to have a deviation from an equal proportion in their sex ratio.

To determine if there is a relationship between habitat suitability and sex ratio, sex ratios for each site were collapsed into a single value. Habitat suitability data was then extracted from the raster and compared to sex ratio data using a regression.

Phenotype was added to the regression to determine of the inclusion of type A or B would increase the relationship among habitat suitability and sex ratio. The regression was compared to just habitat suitability to determine if the model improved.

After exploring the relationship between sex ratio and habitat suitability, a raster containing habitat suitability from the last glacial maximum was used to determine if the availability of suitable habitat was changing. A change in available habitat could be linked to climate change. Change in habitat was calculated and then displayed visually as a bar graph and spatially.

Using the model developed for current conditions, suitability data from the last glacial maximum was used to estimate historical sex ratios, which were then compared to historic habitat suitability.

## Results

The result of the ANOVA test returned a p-value of 0.00307, indicating that there is a difference in sex ratio among sampling locations. 

```{r chunk2, echo=FALSE}


sex_ratios <- samples %>%
  group_by(Site, Plant) %>%
  summarise(
    total_beetles = n(),
    males = sum(Sex == "Male"),
    females = sum(Sex == "Female"),
    sex_ratio = males / total_beetles
  )

site_sex_ratios <- sex_ratios %>%
  group_by(Site) %>%
  summarise(
    mean_sex_ratio = mean(sex_ratio),
    sd_sex_ratio = sd(sex_ratio),
    n = n()
  )

anova_result <- aov(sex_ratio ~ Site, data = sex_ratios)
summary(anova_result)

sex_ratios %>%
  ggplot(aes(x = Site, y = sex_ratio)) +
  geom_boxplot() +
  labs(
    title = "Sex Ratios of Beetles by Site",
    x = "Site",
    y = "Sex Ratio (Proportion of Males)"
  ) +
  theme_minimal()


print("ANOVA Results:")
print(summary(anova_result))

sex_ratio_table <- site_sex_ratios %>%
  # Round the relevant columns to 3 decimal places
  mutate(
    mean_sex_ratio = round(mean_sex_ratio, 3),
    sd_sex_ratio = round(sd_sex_ratio, 3)
  ) %>%
 
  knitr::kable(
    caption = "Summary of Sex Ratios by Site",
    col.names = c("Site", "Mean Sex Ratio", "SD of Sex Ratio", "Number of Plants"),
    align = "c"
  ) %>%
  # Add kable styling options
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )

sex_ratio_table



```

Of the 31 sites sampled, six sites had sex ratios that significantly differed from having equal distribution between male and female.In five of the six sites, males were the dominant sex sampled. 

```{r chunk3, echo=FALSE}

samples <- samples %>%
  mutate(Sex = str_trim(Sex))  

site_sex_counts <- samples %>%
  group_by(Site) %>%
  summarise(
    total_males = sum(Sex == "Male", na.rm = TRUE),
    total_females = sum(Sex == "Female", na.rm = TRUE),
    total_beetles = total_males + total_females
  ) %>%
  mutate(observed_ratio = total_males / total_beetles)

sex_ratio_tests <- site_sex_counts %>%
  rowwise() %>%
  mutate(
    p_value = ifelse(
      total_beetles > 0,
      binom.test(total_males, total_beetles, p = 0.5)$p.value,
      NA_real_
    )
  ) %>%
  ungroup() %>%
  mutate(significant = p_value < 0.05)

sex_ratio_test_table <- sex_ratio_tests %>%
  mutate(
    observed_ratio = round(observed_ratio, 3),
    p_value = round(p_value, 3)
  ) %>%
  knitr::kable(
    caption = "Binomial Test Results for Sex Ratios by Site",
    col.names = c("Site", "Total Males", "Total Females", "Total Beetles", "Observed Ratio", "P-Value", "Significant"),
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )


sex_ratio_test_table



```

Based on the results of the regression, only 2.74% of the variance in sex ratio could be explained by suitability. Additionally, the p-value was 0.3736, indicating that the relationship is not significant. Based on the results of this model, habitat suitability alone is not significant enough to explain sex ratios at the different sites. 

```{r chunk4, echo=FALSE}


arapat_samples <- samples %>%
  mutate(Sex = str_trim(Sex)) 

collapsed_sex_ratios <- arapat_samples %>%
  group_by(Site) %>%
  summarise(
    total_males = sum(Sex == "Male", na.rm = TRUE),
    total_females = sum(Sex == "Female", na.rm = TRUE),
    total_beetles = total_males + total_females,
    observed_ratio = total_males / total_beetles
  ) 

  arapat_locations <- locations %>%
  mutate(Site_number = as.numeric(str_extract(Site, "\\d+"))) 

arapat_locations <- arapat_locations %>%
  arrange(Site_number)

arapat_locations_sf <- st_as_sf(arapat_locations, coords = c("Longitude", "Latitude"), crs = 4326)

overlay_suitability <- raster::extract(suitability_lgm, arapat_locations_sf)
arapat_locations <- arapat_locations %>% mutate(suitability = overlay_suitability)

suitability_sex_ratio <- collapsed_sex_ratios %>%
  left_join(arapat_locations, by = "Site") %>%
  filter(!is.na(suitability))  

model <- lm(observed_ratio ~ suitability, data = suitability_sex_ratio)
regression_summary <- summary(model)

print("Linear Regression Results:")
print(regression_summary)

suitability_sex_ratio %>%
  ggplot(aes(x = suitability, y = observed_ratio)) +
  geom_point(size = 3, color = "blue", alpha = 0.7) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Relationship Between Habitat Suitability and Sex Ratio",
    x = "Habitat Suitability",
    y = "Observed Sex Ratio (Proportion of Males)"
  ) +
  theme_minimal()

site_data_table <- suitability_sex_ratio %>%
  mutate(
    observed_ratio = round(observed_ratio, 3),  # Round observed ratios to 3 decimal places
    suitability = round(suitability, 3)         # Round suitability values to 3 decimal places
  ) %>%
  dplyr::select(Site, observed_ratio, suitability) %>%
  knitr::kable(
    caption = "Summary of Sex Ratios and Habitat Suitability by Site",
    col.names = c("Site", "Observed Sex Ratio", "Habitat Suitability"),
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )

# To display the table in markdown output:
site_data_table


```






```{r chunk5, echo=FALSE}


arapat_locations_sf <- st_as_sf(arapat_locations, coords = c("Longitude", "Latitude"), crs = 4326)

overlay_suitability <- raster::extract(suitability_lgm, arapat_locations_sf)
arapat_locations <- arapat_locations %>% mutate(suitability = overlay_suitability)

raster_df <- as.data.frame(suitability_lgm, xy = TRUE)
colnames(raster_df) <- c("Longitude", "Latitude", "Suitability")


ggplot() +
  geom_raster(data = raster_df, aes(x = Longitude, y = Latitude, fill = Suitability)) +
  scale_fill_gradient(low = "white", high = "red") +  
  geom_sf(data = arapat_locations_sf, color = "black", size = 3) + 
  geom_sf(data = arapat_locations_sf, aes(color = suitability), size = 2) +  
  scale_color_gradient(low = "white", high = "red") +
  labs(
    title = "Habitat Suitability and Location Data",
    x = "Longitude",
    y = "Latitude",
    color = "Habitat Suitability"
  ) +
  theme_minimal() +
  theme(legend.position = "right")



```

By adding phenotype to the suitability model, the model improved slightly in fit. The R squared was 0.02739, with 6.4% of variance in data being explained by the model. Phenotype A had a small positive effect, with a p value of 0.0407, and Phenotype B had a strong negative effect with a p value much less than 0.05. While there was improvement in the model, it was fairly small. 

```{r}
#| echo: false

merged_data <- left_join(suitability_sex_ratio, samples, by = "Site")


# suitability alone
model_1 <- lm(observed_ratio ~ suitability, data = merged_data)
summary(model_1)

# Phenotype_A, and Phenotype_B
model_2 <- lm(observed_ratio ~ suitability + PhenotypeA + PhenotypeB, data = merged_data)
summary(model_2)

cat("Model 1 R-squared:", summary(model_1)$r.squared, "\n")
cat("Model 1 Adjusted R-squared:", summary(model_1)$adj.r.squared, "\n")
cat("Model 2 R-squared:", summary(model_2)$r.squared, "\n")
cat("Model 2 Adjusted R-squared:", summary(model_2)$adj.r.squared, "\n")

cat("Model 1 AIC:", AIC(model_1), "\n")
cat("Model 2 AIC:", AIC(model_2), "\n")

cat("Model 1 p-values:", summary(model_1)$coefficients[,4], "\n")
cat("Model 2 p-values:", summary(model_2)$coefficients[,4], "\n")

anova(model_1, model_2)


```

Based on historic suitability data and current suitability, there have been changes across the different sites and their suitability. Based on the bar graph, some sites experiences a decrease in suitability, while others experienced a slight increase. The maps of the current suitability and the suitability during the glacial maximum show that suitability in the southern region has decreased.Overall, the presence of suitable habitat is decreasing. 

```{r}
#| echo: false

suitability_now_values <- raster::extract(suitability_now, arapat_locations_sf)
suitability_lgm_values <- raster::extract(suitability_lgm, arapat_locations_sf)

arapat_locations <- arapat_locations %>%
  mutate(suitability_now = suitability_now_values,
         suitability_lgm = suitability_lgm_values)

arapat_locations <- arapat_locations %>%
  mutate(suitability_change = suitability_now - suitability_lgm)

ggplot(arapat_locations, aes(x = Site, y = suitability_change)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Change in Habitat Suitability from Last Glacial Maximum to Present",
    x = "Site",
    y = "Change in Suitability (Present - LGM)"
  ) +
  theme_minimal()

suitability_now_df <- as.data.frame(suitability_now, xy = TRUE)
colnames(suitability_now_df) <- c("Longitude", "Latitude", "Suitability_Now")

suitability_lgm_df <- as.data.frame(suitability_lgm, xy = TRUE)
colnames(suitability_lgm_df) <- c("Longitude", "Latitude", "Suitability_LGM")

map_now <- ggplot() +
  geom_raster(data = suitability_now_df, aes(x = Longitude, y = Latitude, fill = Suitability_Now)) +
  scale_fill_gradient(low = "white", high = "red", name = "Suitability") +
  geom_sf(data = arapat_locations_sf, color = "black", size = 1) +
  labs(
    title = "Current Habitat Suitability",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()

map_lgm <- ggplot() +
  geom_raster(data = suitability_lgm_df, aes(x = Longitude, y = Latitude, fill = Suitability_LGM)) +
  scale_fill_gradient(low = "white", high = "red", name = "Suitability") +
  geom_sf(data = arapat_locations_sf, color = "black", size = 1) +
  labs(
    title = "Habitat Suitability During Last Glacial Maximum",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()

print(map_now)
print(map_lgm)

```


Based on the results of the model, as habitat suitability increases, the predicted sex ratio decreases. The negative relationship had a p value of 0.03332, making is significant. These results could have long term effects on the sex ratio of the beetle population as habitat suitability is likely to decrease with the persistance of climate change. 

```{r}
#| echo: false


suitability_now <- raster("Suitability_now.tif")  
suitability_lgm <- raster("Suitability_lgm.asc")  


arapat_locations <- read.csv("Arapat_Locations.csv")   


arapat_locations_sf <- st_as_sf(
  arapat_locations,
  coords = c("Longitude", "Latitude"),
  crs = 4326
)


arapat_locations_sf <- st_transform(arapat_locations_sf, crs = crs(suitability_now))


arapat_locations_sp <- as(arapat_locations_sf, "Spatial")


arapat_locations <- arapat_locations %>%
  mutate(
    current_suitability = raster::extract(suitability_now, arapat_locations_sp),
    lgm_suitability = raster::extract(suitability_lgm, arapat_locations_sp)
  )


arapat_samples <- read.csv("Arapat_Samples.csv")  


observed_sex_ratio <- arapat_samples %>%
  group_by(Site) %>%
  summarise(
    total_males = sum(Sex == "Male", na.rm = TRUE),
    total_females = sum(Sex == "Female", na.rm = TRUE),
    observed_ratio = total_males / total_females
  )


merged_data <- left_join(arapat_locations, observed_sex_ratio, by = "Site")


model_current <- lm(observed_ratio ~ current_suitability, data = merged_data)
summary(model_current)


merged_data <- merged_data %>%
  mutate(predicted_sex_ratio_lgm = predict(model_current, newdata = merged_data %>% mutate(current_suitability = lgm_suitability)))


ggplot(merged_data, aes(x = Site, y = predicted_sex_ratio_lgm)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  labs(
    title = "Predicted Historical Sex Ratio During the Last Glacial Maximum",
    x = "Site",
    y = "Predicted Sex Ratio"
  ) +
  theme_minimal()


ggplot(merged_data, aes(x = lgm_suitability, y = predicted_sex_ratio_lgm)) +
  geom_point(color = "steelblue") +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Predicted Sex Ratio vs. Habitat Suitability (LGM)",
    x = "Habitat Suitability (LGM)",
    y = "Predicted Sex Ratio"
  ) +
  theme_minimal()



aic_current <- AIC(model_current)
print(paste("AIC for current suitability model:", aic_current))

#I know a lot of my code is redundant making it really messy but it is the only way I can stay focused and not get lost in the sauce 

```


