---
title: "Raster"
author: "Ashley Berniche"
format: html
editor: visual
---

#Raster Homework

Load the raster and point data in and crop to an appropriate size to display the locations of the sampling plots and make a label for each site.

```{r}
library(tidyverse)
library(raster)
library(sf)

url <- "https://github.com/DyerlabTeaching/Raster-Data/raw/main/data/alt_22.tif"
raster <- raster(url)

# Load and format beetle data
beetle_url <- "https://raw.githubusercontent.com/dyerlab/ENVS-Lectures/master/data/Araptus_Disperal_Bias.csv"

read_csv(beetle_url) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) -> beetles

beetles %>% st_bbox()

baja_extent <- extent( c(-116, -109, 22, 30 ) )
baja_extent



alt <- crop( raster, baja_extent )

alt.df <- alt %>%
  rasterToPoints() %>%
  as.data.frame() %>% 
  transmute(Longitude = x,
            Latitude = y,
            Elevation = alt_22)

plot(alt)
plot( beetles["Suitability"], pch=16, add=TRUE) +
  text(st_coordinates(beetles)[,1], st_coordinates(beetles)[,2], labels = beetles$Site, pos = 4, cex = 0.7, col = "blue")
```

Use the click() function to crop the raster and filter the sites to include only the sampling locations that are in the Cape regions (e.g., south of site labeled #75).  

```{r}
library(raster)




cropextent <- extent( c(-114, -109, 22, 25 ) )
newmap <- crop( raster, cropextent )
plot(newmap)

beetles <- beetles %>%   
  mutate(Longitude = st_coordinates(.)[, 1],         
         Latitude = st_coordinates(.)[, 2]) 

```

Plot the raster of elevation and all cape region locales with sex-ratio represented as the size of the plot symbol.

```{r}
library(tidyverse)
library(raster)
library(sf)
library(ggplot2)
library(ggrepel)



raster_url <- "https://github.com/DyerlabTeaching/Raster-Data/raw/main/data/alt_22.tif"
cape_raster <- raster(raster_url)  # Load raster data


cape_extent <- extent(c(-111.75, -109.5, 23.25, 24.5))
cape_raster <- crop(cape_raster, cape_extent) 

cape_raster.df <- cape_raster %>%
  rasterToPoints() %>%
  as.data.frame() %>%
  transmute(Longitude = x,
            Latitude = y,
            Elevation = alt_22)  
cape_elevation <- cape_raster.df %>%
  ggplot() + 
  geom_raster(aes(x = Longitude, y = Latitude, fill = Elevation)) + 
  coord_equal() +
  theme_minimal()


site75_coords <- c(Longitude = -110.7460, Latitude = 24.58843)


beetles <- beetles %>%
  mutate(Longitude = st_coordinates(.)[, 1],
         Latitude = st_coordinates(.)[, 2])


cape_sites <- beetles %>%
  filter(Latitude < site75_coords["Latitude"])


cape_map <- cape_elevation + 
  scale_fill_gradient2(low = "darkgreen", mid = "yellow", high = "brown", midpoint = 1000) +
  geom_sf(data = cape_sites, aes(size = MFRatio), color = "dodgerblue2", alpha = 0.75) +
  ggtitle("MF Ratio in Cape Region") +
  labs(x = "Longitude", y = "Latitude") +
  theme_minimal() +
  geom_text_repel(data = cape_sites %>%
                    mutate(Longitude = st_coordinates(.)[, 1],
                           Latitude = st_coordinates(.)[, 2]),
                  aes(x = Longitude, y = Latitude, label = Site),
                  size = 4, color = "black") 

  
 
cape_map

```

The peninsula of Baja California is divided into the States of Baja California Norte and Baja California Sur. The border between these states is at 28° Latitude. Divide the sample locations into groups based on which state they are located in and plot the average sex ratio of the sample sites partitioned by each site.

```{r}

state_sex_ratios <- beetles %>%
  mutate(State = if_else(Latitude >= 28, "Baja California Norte", "Baja California Sur")) %>%
  group_by(State, Site) %>%
  summarize(Avg_MFRatio = mean(MFRatio, na.rm = TRUE)) %>%
  ungroup()

sex_ratio_plot <- state_sex_ratios %>%
  ggplot(aes(x = factor(Site), y = Avg_MFRatio, fill = State)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Site", y = "Average Male-Female Ratio (MFRatio)", 
       title = "Average Sex Ratio by Site and State",
       fill = "State") +
  theme_minimal() +
  scale_fill_manual(values = c("Baja California Norte" = "skyblue", 
                               "Baja California Sur" = "salmon"))

sex_ratio_plot


```



Is there a relationship between the observed sex ratio and the elevation at that site?  Plot these, and in the text, perform a correlation test (cor.test()) and report the values as if you were writing them up in a results section.

```{r}

beetles2 <- beetles %>%
  mutate(Longitude = st_coordinates(.)[, 1],
         Latitude = st_coordinates(.)[, 2]) %>%
  rowwise() %>%
  mutate(Elevation = extract(cape_raster, matrix(c(Longitude, Latitude), ncol = 2))) %>%
  ungroup()

# Plot Sex Ratio vs. Elevation
sex_ratio_elevation_plot <- ggplot(beetles2, aes(x = Elevation, y = MFRatio)) +
  geom_point(color = "dodgerblue") +
  geom_smooth(method = "lm", color = "darkorange", se = TRUE) +
  labs(x = "Elevation (meters)", y = "Male-Female Ratio (MFRatio)", 
       title = "Relationship Between Sex Ratio and Elevation") +
  theme_minimal()

# Display plot
sex_ratio_elevation_plot

# Perform correlation test
correlation_result <- cor.test(beetles2$Elevation, beetles2$MFRatio, use = "complete.obs")

# View correlation test results
correlation_result

```

A Pearson's test was done to determine if there was a relationship between sex ratio and site elevation. The p value was determined to be 0.6508, which far exceeds the accepted significance level. There is insignificant evidence to support a relationship between sex ratio in beetles and site elevation. 