---
title: "TextHomework"
author: "Ashley"
format: html
editor: visual
---

## Text Data Homework

The data for this assignment is found in a text document here.  These data are a textual dump of all course descriptions from Bioinformatics (BNFO), Environmental Studies (ENVS), and Life Sciences (LFSC), in the same format as demonstrated in class.

1. A plot of the number of courses at the 100, 200, 300, and 400-level grouped by program.  


```{r}
library(tidyverse)
library(readr)
library(stringr)


data <- read_lines("C:\\Users\\abern\\OneDrive\\Documents\\ENVS543\\LifeSciencesCurriculum.txt")

class( data )

course_data <- tibble(line = data) %>%
  filter(str_detect(line, "^(ENVS|BNFO|LFSC)\\s+\\d{3}")) %>%
  mutate(
    program = str_extract(line, "^(ENVS|BNFO|LFSC)"),         
    course_number = str_extract(line, "\\d{3}"),             
    level = paste0(str_sub(course_number, 1, 1), "00")        
  )


course_counts <- course_data %>%
  group_by(program, level) %>%
  summarise(course_count = n(), .groups = "drop")


ggplot(course_counts, aes(x = level, y = course_count, fill = program)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Courses by Level and Program",
       x = "Course Level",
       y = "Number of Courses") +
  theme_minimal()




```

2. A wordcloud based upon all course titles.  

```{r}

library(tidyverse)
library(readr)
library(stringr)
library(wordcloud)
library(tm)


data <- read_lines("C:\\Users\\abern\\OneDrive\\Documents\\ENVS543\\LifeSciencesCurriculum.txt")  # Path to the file


course_descriptions <- tibble(line = data) %>%
  filter(!str_detect(line, "^(ENVS|BNFO|LFSC)\\s+\\d{3}")) %>%
  summarise(text = paste(line, collapse = " "))


corpus <- Corpus(VectorSource(course_descriptions$text))


corpus_clean <- corpus %>%
  tm_map(content_transformer(tolower)) %>%              
  tm_map(removePunctuation) %>%                         
  tm_map(removeNumbers) %>%                             
  tm_map(removeWords, stopwords("en")) %>%              
  tm_map(stripWhitespace)                               


dtm <- TermDocumentMatrix(corpus_clean)
matrix <- as.matrix(dtm)
word_freqs <- sort(rowSums(matrix), decreasing = TRUE)   # Sum up word frequencies
df <- data.frame(word = names(word_freqs), freq = word_freqs)


set.seed(123)  # For reproducibility
wordcloud(words = df$word, freq = df$freq, min.freq = 3,
          max.words = 50, random.order = FALSE, rot.per = 0.35,
          colors = brewer.pal(8, "Dark2"))



```

